Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-06 ms/ms.tex

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Effects of Epidemiological Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
%% short title: Epidemiological Structure and HIV Virulence Evolution
%% (53/70 characters)
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Sang Woo Park\textsuperscript{1}
Benjamin M. Bolker\textsuperscript{1,2,3,*}
\\
\bigskip
\textbf{1} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

@@

Pathogens such as HIV can evolve rapidly when the environment changes.
Such changes include decreases in the probability that an infectious
contact (such as a sneeze for a respiratory disease, or an unprotected
sex act for a sexually transmitted disease) encounters an uninfected,
susceptible person and thus has a chance to transmit the
pathogen. Disease spread during an epidemic reduces the fraction of
susceptible people in the population and thus changes the direction of
parasite evolution.  While researchers have explored these
evolutionary dynamics with simulations, their models have neglected
important processes such as people entering and leaving sexual
partnerships. We compared several evolutionary models that include
partnership dynamics and sexual contact outside of stable
partnerships. Models of intermediate complexity predicted lower
virulence over the course of the epidemic (a minimum of 15 years to
progress to AIDS) than either more realistic models or simple models
without any partnership structure at all (both approx. 7.25 years to
progress to AIDS), because sexual contact outside of stable
partnerships tended to wash out the effects of epidemiological
structure. The large effect of epidemiological details on evolutionary
dynamics suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined here as the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
the transmission and disease-induced mortality rate are both linked to the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was 
both heritable and covaried with the set-point viral load (SPVL: i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection) and the probability of transmission.
Higher viral loads led to shorter progression times (higher virulence)
and faster transmission, with a decelerating relationship
between virulence and transmission as required by the tradeoff theory
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
However, despite these studies' attention to detail at the individual
or physiological level, the epidemiological structures used in these
models are relatively simple.

As we discuss in detail below,
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
epidemiological structure in eco-evolutionary models.

We add complexity to the epidemiological model following the general
approach of Champredon \etal\ \cite{champredon_hiv_2013},
which is in turn based on work of Dietz and Hadeler
\cite{dietz_epidemiological_1988}; individuals
join and leave partnerships at a specified rate, and can have sexual
contact both within and outside of established partnerships. In order
to explore how virulence evolution depends on epidemiological
structure, we consider a series of models with increasing levels of
complexity. In order to avoid dependence of the results on a
particular set of parameters --- as we explain below, finding matching
sets of parameters across models with widely differing epidemiological
structures is challenging --- we evaluate our models across a wide
range of parameters, again following Champredon
\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
design. For each model run, we compute a set of metrics (minimum
progression time/peak virulence, timing of maximum virulence, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

As our primary goal is to explore how different epidemiological
structures (i.e. partnership dynamics and contact structures) affect
our conclusions about the evolution of virulence, our models use a
simplified description of within-host dynamics and heritability
derived from Shirreff \etal's multi-strain evolutionary model
\cite{shirreff_transmission_2011}. Like Shirreff \etal, we use a
simple susceptible-infected-susceptible demographic formulation;
rather than modeling birth and death (or more specifically,
recruitment into the sexually active population and death), we assume
that whenever an individual dies from infection, another enters the
susceptible compartment.

@@

We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation; (2) zero vs. positive extra-partnership sexual contact and transmission; and (3) homogeneous vs. heterogeneous levels of sexual activity on the evolution of mean \Lspvl.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and the complexity of epidemiological structures
we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, putting us back where we started when used a naive random-mixing
contact model.
However, we are not quite back where we started, as the
complex models lead to wider, presumably more realistic
confidence intervals on the predictions.
In general, unstructured mixing --- whether occurring through 
purely random mixing, or through extra-pair contact and contact
among people outside of stable partnerships --- tends to drive
faster virulence evolution, leading to higher peak virulence and 
lower times to progression at the peak time.

@@

Parameterization is one of the biggest challenges of epidemiological
modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
by doing Latin hypercube
sampling across a wide range of epidemiological parameters, we 
calibrated each set of parameters to the same initial epidemic
growth rate, chosen to match the results of previous models
\cite{shirreff_transmission_2011}.  Previous models 
in this area have drawn their
parameters from cohort studies from the 1990s
\cite{wawer2005rates,hollingsworth_hiv1_2008}
rather than doing any explicit calibration to epidemic curves,
but they give reasonable order-of-magnitude
growth rates ($\approx 0.04~\textrm{year}^{-1}$)
for the early stages of the HIV epidemic (although considerably
lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
based on population genetic reconstructions \cite{faria_early_2014}).
However, our reason for calibrating was not to match any
specific observed epidemic, but rather to make sure that
we were making meaningful comparisons across a range of
models with radically different epidemiological structures, and
hence involving different interpretations of the same quantitative
parameters.  For example, in models with instantaneous switching the
partnership dissolution rate $c$ is identical to the partnership
formation rate; in models with explicit partnership formation,
the partnership formation rate is also $c$ at equilibrium,
but might vary over the course of an epidemic.
It is not obvious whether models with equal parameters but
different structures should be directly compared; calibration
solves this problem.

@@

Researchers have documented that HIV virulence and set-point viral
load are changing, on time scales comparable to those portrayed here
(e.g., compare \figurename~\ref{fig:virtraj} to Herbeck \etal's
estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
  3] \cite{herbeck_is_2012}), and have begun to build relatively realistic models that
attempt to describe how interventions such as mass antiretroviral
therapy (ART) can be expected to change the trajectory of virulence
evolution \cite{payne_impact_2014,roberts2015impact,herbeck_evolution_2016}.  While these
efforts are well-intentioned, we caution that epidemiological and
other structural details that are currently omitted from these models
could significantly change their conclusions.

\section*{Acknowledgements}
We would like to thank Christophe Fraser and
David Champredon for access to simulation code;
this work was funded by NSERC Discovery Grant 386590-2010.
Computational resources were provided by SHARCnet.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Effects of Contact Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
%% short title: Epidemiological Structure and HIV Virulence Evolution
%% (53/70 characters)
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Sang Woo Park\textsuperscript{1}
Benjamin M. Bolker\textsuperscript{1,2,3,*}
\\
\bigskip
\textbf{1} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

@@

Pathogens such as HIV can evolve rapidly when the environment changes.
Such changes include decreases in the probability that an infectious
contact (such as a sneeze for a respiratory disease, or an unprotected
sex act for a sexually transmitted disease) encounters an uninfected,
susceptible person and thus has a chance to transmit the
pathogen. Disease spread during an epidemic reduces the fraction of
susceptible people in the population and thus changes the direction of
parasite evolution.  While researchers have explored these
evolutionary dynamics with simulations, their models have neglected
important processes such as people entering and leaving sexual
partnerships. We compared several evolutionary models that include
partnership dynamics and sexual contact outside of stable
partnerships. Models of intermediate complexity predicted lower
virulence over the course of the epidemic (a minimum of 15 years to
progress to AIDS) than either more realistic models or simple models
without any partnership structure at all (both approx. 7.25 years to
progress to AIDS), because sexual contact outside of stable
partnerships tended to wash out the effects of contact
structure. The large effect of epidemiological details on evolutionary
dynamics suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
pathogens with higher reproduction numbers ($\rzero$) --- that is, the number of secondary infections caused by a single infected host over the course of its infectious period --- will tend to increase in prevalence relative to strains with lower reproductive ratios.
Pathogens can increase their reproduction numbers either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
transmission and disease-induced mortality rates are both driven by the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.

Fraser \etal\ have applied these ideas to HIV,
showing that the virus appears to satisfy the prerequisites of
the tradeoff theory.
The \emph{set-point viral load} (SPVL: i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection) is a measurable proxy for the rate of HIV within-host
reproduction. Higher viral loads are correlated with 
shorter progression times (higher virulence).
Studies of discordant couples --- stable
sexual partnerships with one infected and one uninfected partner --- %
have shown that SPVL is (1) correlated with faster transmission
(the time to infection of the uninfected partner is shorter when
their partner has a higher SPVL) and (2) heritable
(when the uninfected member of
the couple does become infected, their SPVL is similar to their
partner's).
Furthermore, the rate of transmission increases with
increasing SPVL (and hence with decreasing progression time)
but with a decreasing slope as progression time decreases,
fulfilling the requirements of the tradeoff theory
 \cite{Fraser+2007}.
Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
However, despite these studies' attention to detail at the individual
or physiological level, the contact structures used in these
models are relatively simple.

As we discuss in detail below,
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact %
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
contact structure in eco-evolutionary models.

Because our main goal is to explore how conclusions about virulence
evolution depend on the way in which contact structure
is modeled, we consider a series of
models with increasing levels of complexity in the contact
structure, but simplify some of the other epidemiological
processes (such as the within-host life history of HIV). 
We evaluate our models
across a wide range of parameters, using a Latin hypercube design; for
each model run, we compute a set of metrics 
that summarize the evolutionary trajectory of SPVL
over the course of the epidemic.

@@

We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation; (2) zero vs. positive extra-partnership sexual contact and transmission; and (3) homogeneous vs. heterogeneous levels of sexual activity on the evolution of mean \Lspvl.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and the complexity of contact structures
we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, putting us back where we started when used a naive random-mixing
contact model.
However, we are not quite back where we started, as the
complex models lead to wider, presumably more realistic
confidence intervals on the predictions.
In general, unstructured mixing --- whether occurring through 
purely random mixing, or through extra-pair contact and contact
among people outside of stable partnerships --- tends to drive
faster virulence evolution, leading to higher peak virulence and 
lower times to progression at the peak time.

@@

Parameterization is one of the biggest challenges of epidemiological
modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
by doing Latin hypercube
sampling across a wide range of epidemiological parameters, we 
calibrated each set of parameters to the same initial epidemic
growth rate, chosen to match the results of previous models
\cite{shirreff_transmission_2011}.  Previous models 
in this area have drawn their
parameters from cohort studies from the 1990s
\cite{wawer2005rates,hollingsworth_hiv1_2008}
rather than doing any explicit calibration to epidemic curves,
but they give reasonable order-of-magnitude
growth rates ($\approx 0.04~\textrm{year}^{-1}$)
for the early stages of the HIV epidemic (although considerably
lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
based on population genetic reconstructions \cite{faria_early_2014}).
However, our reason for calibrating was not to match any
specific observed epidemic, but rather to make sure that
we were making meaningful comparisons across a range of
models with radically different contact structures, and
hence involving different interpretations of the same quantitative
parameters.  For example, in models with instantaneous switching the
partnership dissolution rate $c$ is identical to the partnership
formation rate; in models with explicit partnership formation,
the partnership formation rate is also $c$ at equilibrium,
but might vary over the course of an epidemic.
It is not obvious whether models with equal parameters but
different structures should be directly compared; calibration
solves this problem.

@@

Researchers have documented that HIV virulence and set-point viral
load are changing, on time scales comparable to those portrayed here
(e.g., compare \figurename~\ref{fig:virtraj} to Herbeck \etal's
estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
  3] \cite{herbeck_is_2012}), and have begun to build relatively realistic models that
attempt to describe how interventions such as mass antiretroviral
therapy (ART) can be expected to change the trajectory of virulence
evolution \cite{payne_impact_2014,roberts2015impact,herbeck_evolution_2016}.  While these
efforts are well-intentioned, we caution that 
structural details that are currently omitted from these models
could significantly change their conclusions.

\section*{Acknowledgements}
We would like to thank Christophe Fraser and
David Champredon for access to simulation code
and Wim Delva and an anonymous reviewer for useful comments;
this work was funded by NSERC Discovery Grant 386590-2010.
Computational resources were provided by SHARCnet.
